---
title: "Using R for Analytic Graphs: Learn How Data Visualization Can Improve Interpretation in Social Work Research"
date: "Saturday, November 01, 2014"
output: beamer_presentation
---


# Why Use R?

- Free

- Open Source

- Easy Collaboration

- Replicable Research 

# Why Wouldn't You Use R?

Steep(er) learning curve compared to, say, Excel or SPSS. This matters a lot if

- You run statistics rarely.

- You want a point and click interface. 

# Where Can you Get R?

- CRAN

- Our Thumb Drives

# Where Are We Going Today?

- Graphing Model Results

- Graphing Other Things

# Graphing Model Results (Basic Algorithm)

1. Choose a counterfactual $x_c$.

2. Estimate a model to get a vector of parameters $\hat{\boldsymbol\beta}$ and the associated variance-covariance matrix, $\hat{\boldsymbol V}$.

3. Draw several $\tilde{\boldsymbol\beta}$ from $\mathcal{N}(\hat{\boldsymbol\beta},\,\hat{\boldsymbol V})$, where $\mathcal{N}$ is a mulivariate normal distribution. 

4. Calculate expected outcomes based on model parameters for all of your draws from $\mathcal{N}$. 

5. Calculate summary statistics for each level of $x_c$.

This approach will work for most of the models that social welfare researchers tend to encounter.

# A Practical Example - Background

## Research Question

How does a child's probability of exiting the foster care system vary by child characteristics?

## Multiple Permanency Outcomes

Requires that we estimate a mulinomial logistic regression model.

## Data in Question 

- 619 Children entering out-of-home care in late 2007. 

- Children's parent's were surveyed once in 2007. The survey results were then linked to administrative data which faciliated a longitudinal follow-up. 

NOTE: I will simulate a data set for sharing. I cannot share the raw data. 

# A practical example - Choose a counterfactual $x_c$.

## Load the data 

```{r}
dat <- read.csv("dat.csv")
```

# A practical example - Choose a counterfactual $x_c$.

```{r, fig.width=3, fig.height=2, warning=FALSE, message=FALSE}
#looking at age of child at episode begin
require(ggplot2)
ggplot(dat, aes(x=age_eps_begin)) + 
  geom_histogram(binwidth = 1)
```

# A practical example - Choose a counterfactual $x_c$.

```{r, fig.width=6, fig.height=3, warning=FALSE, message=FALSE}
#looking at age of child at episode begin by outcome 
ggplot(dat, aes(x=age_eps_begin, fill=outcome)) + 
  geom_histogram(binwidth = 1) +
  facet_wrap(~ outcome)
```

# A practical example - Estimate a model. 

## Need to estimate a statistical model to get 

1. A vector of parameters $\hat{\boldsymbol\beta}$, and 

2. The associated variance-covariance matrix, $\hat{\boldsymbol V}$.

# A practical example - Estimate a model. 

## Prep the data

```{r, fig.width=6, fig.height=3, warning=FALSE, message=FALSE}
# easy to load external packages
# install.packages("nnet") # install once
require(nnet)             # load every time

# recode data
levels(dat$outcome) <- c("Adoption", "Emancipation"
                         ,"Guardianship", "Reunification")

# relevel our outcome variable
dat$outcome_rl <- relevel(dat$outcome
                          , ref = "Emancipation")

# recode to numeric
dat$outcome_rl <- as.numeric(dat$outcome_rl)
```

# A practical example - Estimate a model. 

## Run the model

```{r, warning=FALSE, message=FALSE}
# run the multinomial model
model <- multinom(outcome_rl ~ age_eps_begin + 
                   eps_rank 
                 ,data = dat
                 ,Hess = TRUE)
```


# A practical example - Estimate a model. 

## Display of summary the model

```{r, warning=FALSE, message=FALSE, }
model
```


# A practical example - Estimate a model. 

## Extract a vector of parameters $\hat{\boldsymbol\beta}$

```{r, warning=FALSE, message=FALSE}
#run the multinomial model
pe <- model$wts[c(6,7,8,10,11,12,14,15,16)]
pe[1:3]
pe[4:6]
pe[7:9]

```

# A practical example - Estimate a model. 

## Extract the associated variance-covariance matrix, $\hat{\boldsymbol V}$

```{r, warning=FALSE, message=FALSE}
#run the multinomial model
vc <- solve(model$Hess) 
```

# A practical example - Draw several $\tilde{\boldsymbol\beta}$ from $\mathcal{N}(\hat{\boldsymbol\beta},\,\hat{\boldsymbol V})$. 

```{r, warning=FALSE, message=FALSE}
#load a package which contains a multivariate normal 
#sampling function
require(MASS)
#assign a variable for the number of simulations
sims <- 10000
#draw the indicates number of beta simulates 
#using our extracted model data
simbetas <- mvrnorm(sims,pe,vc)
```


# A practical example - Last two steps...


- Calculate expected values for all of your draws from $\mathcal{N}$, and

- Calculate summary statistics for each level of $x_c$.

- Specific calculations are beyond the scope of this presentation

- But the `simcf` package from Chris Adolph (political scientist at the University of Washington) will do them for us!


# A practical example - Last two steps

## Get data read for `simcf`

- Re-arrange simulates to array format

```{r, warning=FALSE, message=FALSE}
simb <- array(NA, dim = c(sims,3,3))
simb[,,1] <- simbetas[,1:3]         
simb[,,2] <- simbetas[,4:6]
simb[,,3] <- simbetas[,7:9]
```

- Specify range of counterfactual values

```{r, warning=FALSE, message=FALSE}
agerange <- seq(0,17,by=0.1)    
```

# A practical example - Last two steps

## Get data read for `simcf`

- Load `simcf` and use the `cfFactorial()` function to set specific values for simulation. 

```{r, warning=FALSE, message=FALSE}
require(simcf)
xhyp <- cfFactorial(age = agerange
                    ,ep_rank = mean(dat$eps_rank))
```

- Run the simulation (this is where the last two steps are really performed).

```{r, warning=FALSE, message=FALSE}
test_sims <- mlogitsimev(xhyp,simb,ci=0.95)
```

# Get the data ready to graph

```{r, warning=FALSE, message=FALSE}
y <- as.vector(test_sims$pe[,1:4])

x <- rep(1:length(agerange), 4)

lower <- as.vector(test_sims$lower[,1:4,])

upper <- as.vector(test_sims$upper[,1:4,])

Outcome <- c(rep("Adoption", length(agerange))
                 ,rep("Guardianship"
                      ,length(agerange))
                 ,rep("Reunification"
                      ,length(agerange))
                 ,rep("Emancipation"
                      ,length(agerange)))
```

# Get the data ready to graph

```{r, warning=FALSE, message=FALSE}
dat_sim_plot <- data.frame(y,x,lower,upper,Outcome)
```

# Graph the data!

```{r,  warning=FALSE, message=FALSE}
p1 <- ggplot(dat_sim_plot
       ,aes(x=x/10, y=y, group=Outcome)) + 
        geom_line() 
```

# Graph the data!

```{r, warning=FALSE, message=FALSE, echo=FALSE}
p1
```

# Make it Pretty!

```{r,  warning=FALSE, message=FALSE}
p2 <- ggplot(dat_sim_plot
       ,aes(x=x/10, y=y, group=Outcome)) + 
        geom_line(size=1, alpha=.5) +
        geom_ribbon(aes(ymin=lower
                        ,ymax=upper
                        ,fill=Outcome), alpha=.5) +
        ylab("Pr(Outcome|Age,Prior Episodes)") +
        xlab("Age at Entry into Foster Care") +
        theme_bw() 
```


# Make it Pretty

```{r, warning=FALSE, message=FALSE, echo=FALSE}
p2
```
